---
title: "Bayes modeling"
author: "Megan Sears"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: yes
    theme: journal
editor_options:
  chunk_output_type: console
---

```{r, include = F}

library(pacman)

p_load(tidyverse, 
       reshape, 
       broom, 
       car, 
       ggdist, 
       brms, 
       rstanarm, 
       bayesplot, 
       performance,
       beepr,
       loo)

# Set a basic ggplot theme
theme_set(theme_bw(base_size = 20))

```

# Data prep

```{r}

response <- read_csv('./data/final_model_inputs/events_inputs_v2.csv') %>%
  select(site,
         sp_cat, # categorical
         fire, # categorical
         stage_rise_cm,
         lag2peak_halfP,
         lag2peak_Pstart,
         MI60_mmhr,
         lag_pwd,
         terrain_hypso75 = hypso75,
         terrain_hypso25 = hypso25,
         terrain_hypso50 = hypso50,
         terrain_connec = IC,
         terrain_conn_dnbr = hydraulic_func_conn,
         area_km2 = area_km2.x,
         terrain_rc_circle = RC_CIRCLE.x,
         terrain_slope40 = slope40_frac,
         soil_brockdepmin = brockdepmin,
         terrain_recharge = frac_recharge_area,
         #terrain_vbw = avg_vbw_m,
         soil_sand = AVG_SAND,
         mean_dnbr, 
         soil_clay = AVG_CLAY,
         soil_ksat = AVG_KSAT, 
         terrain_elongation = ELONGATION.x, 
         soil_availwaterstor = aws0150wta, 
         NDVI, 
         terrain_vall_holl = frac_valley,
         terrain_flowlength = flow_length_m, 
         terrain_slope30 = slope30_frac, 
         SP_mean,
         terrain_hypsoint = hypso_integral,
         year,
         terrain_glacial = frac_glacial,
         fire_no,
         mulch_frac) %>%
  mutate(site_number = as.integer(factor(site))) %>% # give each site a number
  select(-site)

# prep df for stage rise
stage <- response %>%
  select(-c(lag2peak_halfP, 
            lag2peak_Pstart,
            fire_no)) %>% # remove lag to peak
  relocate(year, .after = site_number) %>% # move year to end of data set to make next line easier to code
  mutate(across(c(MI60_mmhr:mulch_frac), scale), # scale predictor data to mean = 0, sd =1
         site_number = factor(site_number),
         yearFact = factor(year), # year factor for random effect
         yearCont = year - 2022) # centered year for fixed effect

# prep df for lag to peak
lag <- response %>%
  select(-c(lag2peak_Pstart,
            stage_rise_cm,
            fire_no)) %>%
  filter(!lag2peak_halfP <= 0) %>%
  relocate(year, .after = site_number) %>% # move year to end of data set to make next line easier to code
  mutate(across(c(MI60_mmhr:mulch_frac), scale), # scale predictor data to mean = 0, sd =1
         site_number = factor(site_number),
         yearFact = factor(year), # year factor for random effect
         yearCont = year - 2022) # centered year for fixed effect

```

# Stage response

## Global

- Fit one large model with no interactions, allows differences among watersheds to be defined by random effects
- Informs us if watershed differences need further investigation

**
Those divergence warnings go away if you shrink the priors (prior = normal(0, 1), prior_intercept = normal(0, 1)), but it doesn’t change the model results that much or any of the check_model diagnostics, while taking a little longer to converge so I was using the broader ranges I included in the code I sent you and ignoring the warnings.

From that link:
Even a small number of divergences after warmup cannot be safely ignored if completely reliable inference is desired. But if you get only few divergences and you get good Rhat and ESS values, the resulting posterior is often good enough to move forward. There are also cases when a small number divergences without any pattern in their locations can be verified to be unimportant, but this cannot be safely assumed without a careful investigation of the model.

If you look at the MCMC diagnostics in your model summary, all Rhat values = 1.0 (perfect convergence), and the estimated sample sizes (ESS; ‘n_eff’) are all >20% of the post-warm up iterations. Which is good. But for full defensibility you may just want to shrink your priors to get rid of the error and concern.

If shrinking your priors, you should probably do it in the other models too.
**

```{r}

# global model for stage rise
global <- stan_glmer(
  stage_rise_cm ~ 
    MI60_mmhr + 
    fire + 
    sp_cat + 
    yearCont +       
    (1 + MI60_mmhr | site_number) + 
    (1|yearFact),                          
  data = stage,
  family = Gamma(link = "log"),
  prior = normal(0, 1), # default rec from Gelman et al. for gamma models is 0,2.5
  prior_intercept = normal(0, 1),
  prior_covariance = decov(regularization = 2),
  chains = 5,
  cores = 2,
  iter = 5000,
  adapt_delta = 0.99,
  init = 0
)

summary(global)
check_model(global)
r2(global)

# globalLess <- stan_glmer(
#   stage_rise_cm ~ 
#     MI60_mmhr + 
#     fire + 
#     sp_cat + 
#     yearCont +       
#     (1 | site_number) + 
#     (1|yearFact),                          
#   data = stage,
#   family = Gamma(link = "log"),
#   prior = normal(0, 1), # default rec from Gelman et al. for gamma models is 0,2.5
#   prior_intercept = normal(0, 1),
#   prior_covariance = decov(regularization = 2),
#   chains = 5,
#   cores = 2,
#   iter = 5000,
#   adapt_delta = 0.99,
#   init = 0
# )
# 
# loo_global <- loo(global,k_threshold = 0.7)
# loo_globalLess <- loo(globalLess, k_threshold = 0.7)
# 
# loo_compare(loo_global, loo_globalLess) # global is better
print(loo_global) # k estimates are good

# saveRDS(global, 
#         "./data/bayes_interaction_outputs/global_stageRise_final.rds")

# look at posterior distr and credible intervals
as_draws(global) %>% 
  data.frame() %>%
  select(MI60_mmhr, 
         fireetf, 
         sp_catlow, 
         yearCont) %>%
  melt() %>%
  ggplot(., aes(x = value, 
                fill = variable)) + 
  theme_bw(base_size = 20) +
  geom_vline(xintercept = 0) +
  stat_slab(aes(fill = after_stat(level)), 
            .width = c(.66, .95, 1), 
            color = 'gray50') +
  stat_pointinterval(.width = c(.66, .95, 1), 
                     fatten_point = 2.5, 
                     color = 'darkred', 
                     interval_size_domain = c(2,5,10)) +
  paletteer::scale_fill_paletteer_d("soilpalettes::gley") +
  facet_wrap(~variable, 
             ncol = 1, 
             scales = 'free_y') +
  theme(legend.position = 'none')

```

## NDVI

```{r}

# # include three way interaction between MI60, NDVI, and year to assess those time varying effects
# ndviYear <- stan_glmer(
#   stage_rise_cm ~ 
#     MI60_mmhr*yearCont*NDVI + 
#     fire + 
#     sp_cat +       
#     (1 + MI60_mmhr | site_number) + 
#     (1|yearFact),                          
#   data = stage,
#   family = Gamma(link = "log"),
#   prior = normal(0, 1),
#   prior_intercept = normal(0, 1),
#   prior_covariance = decov(regularization = 2),
#   chains = 5,
#   cores = 8,
#   iter = 5000,
#   adapt_delta = 0.99,
#   init = 0
# )
# 
# summary(ndviYear)
# check_model(ndviYear)
# r2(ndviYear)

# ndvi less
ndviLess <- stan_glmer(
  stage_rise_cm ~ 
    MI60_mmhr*yearCont*NDVI + 
    fire + 
    sp_cat +       
    (1 | site_number) + 
    (1|yearFact),                          
  data = stage,
  family = Gamma(link = "log"),
  prior = normal(0, 1),
  prior_intercept = normal(0, 1),
  prior_covariance = decov(regularization = 2),
  chains = 5,
  cores = 8,
  iter = 5000,
  adapt_delta = 0.99,
  init = 0
)

summary(ndviLess)
check_model(ndviLess)
r2(ndviLess)

# # test using loo cross val, got a k>0.7 warning
# loo_ndviYear <- loo(ndviYear, k_threshold = 0.7)
# loo_ndviLess <- loo(ndviLess, k_threshold = 0.7)
# 
# # ndviYear does not improve the predictive performance and performs slightly worse
# loo_compare(loo_ndviYear, loo_ndviLess)

# loo ndviLess is not overfitting
print(loo_ndviLess)

# as_draws(ndviYear) %>% 
#   data.frame() %>%
#   select(NDVI, 
#          MI60_mmhr.NDVI, 
#          yearCont.NDVI, 
#          MI60_mmhr.yearCont.NDVI) %>%
#   melt() %>%
#   ggplot(., aes(x = value, 
#                 fill = variable)) + 
#   theme_bw(base_size = 20) +
#   geom_vline(xintercept = 0) + 
#   stat_slab(aes(fill = after_stat(level)), 
#             .width = c(.66, .95, 1), 
#             color = 'gray50') +
#   stat_pointinterval(.width = c(.66, .95, 1), 
#                      fatten_point = 2.5, 
#                      color = 'darkred', 
#                      interval_size_domain = c(2,5,10)) +
#   paletteer::scale_fill_paletteer_d("soilpalettes::gley") +
#   facet_wrap(~variable, 
#              ncol = 1) +
#   theme(legend.position = 'none')

saveRDS(ndviLess, 
        "./data/bayes_outputs/ndvi_stageRise_final.rds")

```

## Interaction w/ PCs

Tested a full and reduced bayes model. The reduced scored better based on loo compared, or I wasn't gaining anything with the full. 

```{r}

pca <- read_csv('./data/pca_output/pcExport.csv') %>%
  mutate(
    PC1_scale = as.numeric(scale(PC1)),
    PC2_scale = as.numeric(scale(PC2))) 

# interFull <- stan_glmer(
#   stage_rise_cm ~
#     MI60_mmhr*PC1 +
#     MI60_mmhr*PC2 +
#     yearCont +
#     (1 + MI60_mmhr | site_number) +
#     (1 | yearFact),
#   data = pca,
#   family = Gamma(link = "log"),
#   prior = normal(0, 1), # default rec from Gelman et al. for gamma models is 0,2.5
#   prior_intercept = normal(0, 1),
#   prior_covariance = decov(regularization = 2),
#   chains = 5,
#   cores = 2,
#   iter = 5000,
#   adapt_delta = 0.99,
#   init = 0
# )
# 
# summary(interFull)
# check_model(interFull)
# r2(interFull)

interSR <- stan_glmer(
  stage_rise_cm ~ 
    MI60_mmhr*PC1_scale + 
    MI60_mmhr*PC2_scale +
    yearCont + 
    (1 | site_number) + 
    (1 | yearFact),                   
  data = pca,
  family = Gamma(link = "log"),
  prior = normal(0, 1), # default rec from Gelman et al. for gamma models is 0,2.5
  prior_intercept = normal(0, 1),
  prior_covariance = decov(regularization = 2),
  chains = 5,
  cores = 8,
  iter = 5000,
  adapt_delta = 0.99,
  init = 0
)

summary(interSR)
check_model(interSR)
r2(interSR)

# compare the full and reduced models
# test using loo cross val, got a k>0.7 warning
# loo_interFull <- loo(interFull,k_threshold = 0.7)
# loo_interSR <- loo(interSR,k_threshold = 0.7)
# 
# # interFull does not improve the predictive performance and performs slightly worse
# loo_compare(loo_interFull, loo_interSR)

print(loo_interSR) # k estimates are good

saveRDS(interSR,
        "./data/bayes_outputs/interaction_stageRise_bayes.rds")

srInterPlot <- as_draws(interSR) %>% 
  data.frame() %>%
  select(4,6,7,8) %>% #2,6,7
  melt() %>%
  ggplot(., 
                      aes(x = value,
                          y=variable)) +
  geom_vline(xintercept = 0) +
  stat_slab(aes(fill = after_stat(level)), 
            .width = c(.66, .95, 1),
            color='grey',
            linewidth = 0.3) +
  stat_pointinterval(.width = c(.66, .95, 1), 
                     fatten_point = 2.5,
                     interval_size_domain = c(2,5,10),
                     color = '#D50032FF') +
  paletteer::scale_fill_paletteer_d("nbapalettes::kings_city") +
  #scale_y_discrete(limits = rev(var_levels)) +
    theme(legend.position = 'none',
        strip.background = element_rect(fill = "white", color = NA),
        strip.text = element_text(hjust = 0),
        text = element_text(color = "black")) +
  labs(x = 'Scaled coefficient',
       y = NULL) 

srInterPlot

```

## PWD

```{r}

pwdModel <- stan_glmer(
  stage_rise_cm ~ 
    MI60_mmhr*lag_pwd + 
    fire + 
    sp_cat + 
    yearCont +
    (1 | site_number) + 
    (1|yearFact),                          
  data = stage,
  family = Gamma(link = "log"),
  prior = normal(0, 1),
  prior_intercept = normal(0, 1),
  prior_covariance = decov(regularization = 2),
  chains = 5,
  cores = 8,
  iter = 5000,
  adapt_delta = 0.99,
  init = 0
)

summary(pwdModel)
check_model(pwdModel)
r2(pwdModel)

loo_pwdModel <- loo(pwdModel, k_threshold = 0.7)
print(loo_pwdModel) # k estimates are good

saveRDS(pwdModel,
        "./data/bayes_outputs/pwd_stageRise_bayes.rds")

pwdPlot <- as_draws(pwdModel) %>% 
  data.frame() %>%
  select(2:7) %>%
  melt() %>%
  ggplot(., 
                      aes(x = value,
                          y=variable)) +
  geom_vline(xintercept = 0) +
  stat_slab(aes(fill = after_stat(level)), 
            .width = c(.66, .95, 1),
            color='grey',
            linewidth = 0.3) +
  stat_pointinterval(.width = c(.66, .95, 1), 
                     fatten_point = 2.5,
                     interval_size_domain = c(2,5,10),
                     color = '#D50032FF') +
  paletteer::scale_fill_paletteer_d("nbapalettes::kings_city") +
  #scale_y_discrete(limits = rev(var_levels)) +
    theme(legend.position = 'none',
        strip.background = element_rect(fill = "white", color = NA),
        strip.text = element_text(hjust = 0),
        text = element_text(color = "black")) +
  labs(x = 'Scaled coefficient',
       y = NULL) 

pwdPlot


pwdSummary <- as_draws(pwdModel) %>% 
  data.frame() %>%
  select(2:7) %>%
  melt() %>% 
  group_by(variable) %>%
  summarize(mean = mean(value),
            median = median(value),
            min = min(value),
            max = max(value),
            lower_66 = quantile(value, 0.17),
            upper_66 = quantile(value, 0.83),
            lower_95 = quantile(value, 0.025),
            upper_95 = quantile(value, 0.975),
            percent_positive = round(mean(value > 0) * 100, 1)) %>%
  mutate(percent_positive_label = paste0("P>0 = ", 
                                         sprintf("%.1f", 
                                                 percent_positive), "%"))


```

# Lag to peak

## Global

```{r}

# global model for lag to peak
global <- stan_glmer(
  lag2peak_halfP ~ 
    MI60_mmhr + 
    fire + 
    sp_cat + 
    yearCont +       
    (1 + MI60_mmhr | site_number) + 
    (1|yearFact),                          
  data = lag,
  family = Gamma(link = "log"),
  prior = normal(0, 1), # default rec from Gelman et al. for gamma models is 0,2.5
  prior_intercept = normal(0, 1),
  prior_covariance = decov(regularization = 2),
  chains =5, # switched to 5 when ready for final results
  cores = 2,
  iter = 5000, # switch to 5k when ready for final results
  adapt_delta = 0.99,
  init = 0
)

summary(global)
check_model(global)
r2(global)

loo_globalL2P <- loo(global, k_threshold = 0.7)
print(loo_globalL2P) # estimates are good

saveRDS(global, 
        "./data/bayes_interaction_outputs/global_l2p_final.rds")

# lag to peak posterior draws
as_draws(global) %>% 
  data.frame() %>%
  select(MI60_mmhr, 
         fireetf, 
         sp_catlow, 
         yearCont) %>%
  melt() %>%
  ggplot(., aes(x = value, 
                fill = variable)) + 
  theme_bw(base_size = 20) +
  geom_vline(xintercept = 0) +
  stat_slab(aes(fill = after_stat(level)), 
            .width = c(.66, .95, 1), 
            color = 'gray50') +
  stat_pointinterval(.width = c(.66, .95, 1), 
                     fatten_point = 2.5, color = 'darkred', 
                     interval_size_domain = c(2,5,10)) +
  paletteer::scale_fill_paletteer_d("soilpalettes::gley") +
  facet_wrap(~variable, 
             ncol = 1, 
             scales = 'free_y') +
  theme(legend.position = 'none') 

```

## NDVI

```{r}

# include three way interaction between MI60, NDVI, and year to assess those time varying effects
# changed the random effect for site number to remove MI60 based on loo results in stage rise ndvi model
ndviYear <- stan_glmer(
  lag2peak_halfP ~ 
    MI60_mmhr*yearCont*NDVI + 
    fire + 
    sp_cat +       
    (1|site_number) + 
    (1|yearFact),                          
  data = lag,
  family = Gamma(link = "log"),
  prior = normal(0,1),
  prior_intercept = normal(0, 1),
  prior_covariance = decov(regularization = 2),
  chains = 5,
  cores = 8,
  iter = 5000,
  adapt_delta = 0.99,
  init = 0
)

summary(ndviYear)
check_model(ndviYear)
r2(ndviYear)

l2pNDVIloo <- loo(ndviYear,
                  k_threshold = 0.7)

print(l2pNDVIloo)
# All Pareto k estimates are good (k < 0.7).

saveRDS(ndviYear, 
        "./data/bayes_outputs/ndvi_l2p_final.rds")

# posterior draw distr
as_draws(ndviYear) %>% 
  data.frame() %>%
  select(NDVI, 
         MI60_mmhr.NDVI,
         yearCont.NDVI, 
         MI60_mmhr.yearCont.NDVI) %>%
  melt() %>%
  ggplot(., aes(x = value, 
                fill = variable)) + 
  theme_bw(base_size = 20) +
  geom_vline(xintercept = 0) + 
  stat_slab(aes(fill = after_stat(level)), 
            .width = c(.66, .95, 1), 
            color = 'gray50') +
  stat_pointinterval(.width = c(.66, .95, 1), 
                     fatten_point = 2.5, 
                     color = 'darkred', 
                     interval_size_domain = c(2,5,10)) +
  paletteer::scale_fill_paletteer_d("soilpalettes::gley") +
  facet_wrap(~variable, 
             ncol = 1) +
  theme(legend.position = 'none')

```

## Interaction w/ PCs

```{r}

pca <- read_csv('./data/pca_output/pcExport.csv') %>%
  mutate(
    PC1_scale = as.numeric(scale(PC1)),
    PC2_scale = as.numeric(scale(PC2))) %>%
   filter(!lag2peak_halfP <= 0)
  

interL2P <- stan_glmer(
  lag2peak_halfP ~ 
    MI60_mmhr*PC1_scale + 
    MI60_mmhr*PC2_scale +
  yearCont + (1 | site_number) + 
  (1 | yearFact),                   
  data = pca,
  family = Gamma(link = "log"),
  prior = normal(0, 1), # default rec from Gelman et al. for gamma models is 0,2.5
  prior_intercept = normal(0, 1),
  prior_covariance = decov(regularization = 2),
  chains = 5,
  cores = 2,
  iter = 5000,
  adapt_delta = 0.99,
  init = 0
)

summary(interL2P)
check_model(interL2P)
r2(interL2P)

loo_interL2P <- loo(interL2P, k_threshold = 0.7)
print(loo_interL2P) # estimates are good

saveRDS(interL2P, 
        "./data/bayes_outputs/interaction_lag2peak_bayes.rds")

l2pInterPlot <- as_draws(interL2P) %>% 
  data.frame() %>%
  select(2,6,7) %>%
  melt() %>%
  ggplot(., 
                      aes(x = value,
                          y=variable)) +
  geom_vline(xintercept = 0) +
  stat_slab(aes(fill = after_stat(level)), 
            .width = c(.66, .95, 1),
            color='grey',
            linewidth = 0.3) +
  stat_pointinterval(.width = c(.66, .95, 1), 
                     fatten_point = 2.5,
                     interval_size_domain = c(2,5,10),
                     color = '#D50032FF') +
  paletteer::scale_fill_paletteer_d("nbapalettes::kings_city") +
  #scale_y_discrete(limits = rev(var_levels)) +
    theme(legend.position = 'none',
        strip.background = element_rect(fill = "white", color = NA),
        strip.text = element_text(hjust = 0),
        text = element_text(color = "black")) +
  labs(x = 'Scaled coefficient',
       y = NULL) 

l2pInterPlot

```

## PWD

```{r}

pwdModel <- stan_glmer(
  lag2peak_halfP ~ 
    MI60_mmhr*lag_pwd + 
    fire + 
    sp_cat + 
    yearCont +
    (1 | site_number) + 
    (1|yearFact),                          
  data = lag,
  family = Gamma(link = "log"),
  prior = normal(0, 1),
  prior_intercept = normal(0, 1),
  prior_covariance = decov(regularization = 2),
  chains = 5,
  cores = 8,
  iter = 5000,
  adapt_delta = 0.99,
  init = 0
)

summary(pwdModel)
check_model(pwdModel)
r2(pwdModel)

loo_pwdModel <- loo(pwdModel, k_threshold = 0.7)
print(loo_pwdModel) # k estimates are good

saveRDS(pwdModel,
        "./data/bayes_outputs/pwd_l2p_bayes.rds")

```
